---
output:
  html_document: default
  pdf_document: default
---
# R for Statistical Inference

## Descriptive statistics
When working with real-life data, one of the first things we may want to do is get a general understanding of the data. For this purpose, we can use some descriptive statics and plots available in R. 

R comes with some databases. To see the list of data sets available, type ``data()``. For example, you will find the ``iris`` data set (type ``?iris`` for a complete description of the data). Although not an insurance or financial data set, we will use ``iris`` in this section since it is readily available and the tools presented here do not depend on the type of data.

First, we need to load the data into our working space using ``data()``:
```{r}
data(iris)
```
Now, the ``iris`` should appear in the **Environment** tab. Note that ``iris`` is a data frame, which is one of the most common ways of presenting data in R. 
```{r}
str(iris)
```

::: {.remark}
Some R packages include data sets. For instance, install and load the package ``insuranceData``. Here you will find the ``dataCar`` data set. 
:::

The first thing that we may want to do is to compute some measures of central tendency and variability. Let us focus for now on the column ``Sepal.Lenght``. Regarding measures of central tendency, these refer to measures such as the mean and the median. We already know that the function ``mean()`` can be used to compute the sample mean. 
```{r}
mean(iris$Sepal.Length)
```
We can also compute the sample median using the ``median()`` function. Recall that the sample median refers to the value in the middle of the observations. 
```{r}
median(iris$Sepal.Length)
```
Let us now move to some measures of variability, which help us understand how spread the data is. First, we can look at the minimum and maximum values using the functions ``min()`` and ``max()``, receptively. 
```{r}
min(iris$Sepal.Length)
max(iris$Sepal.Length)
```
Alternatively, we can use the ``range()`` function to compute both minimum and maximum simultaneously. 
```{r}
range(iris$Sepal.Length)
```
Next, we can look at the variance, which measures how the data values are dispersed around the mean. This can be computed with the ``var()`` function: 
```{r}
var(iris$Sepal.Length)
```
A related measure is the standard deviation, which is simply the square root of the variance. To compute it, we can use ``sd()``:
```{r}
sd(iris$Sepal.Length)
```
Finally, we may be interested in computing the sample quantiles of our data. The R command to do so is ``quantile()``:
```{r}
quantile(iris$Sepal.Length)
```
Note that by default, R computes the quantiles of 0% (min), 25%, 50% (median), 75% and 100% (max). We can change this in with the argument ``probs``:
```{r}
quantile(iris$Sepal.Length, probs = seq(0.1, 0.9, by = 0.1))
```

Although all the above measures can be computed separately, the ``summary()`` function provides an easier way to compute several of them at once:
```{r}
summary(iris$Sepal.Length)
```
Moreover, if our data set has several columns, ``summary()`` can compute all these measures for all columns at the same time: 
```{r}
summary(iris)
```


Note that our data set has a column called ``Species`` indicating the flower species. Although the summary function above already provide us with the list of species (three species), we can also use the ``unique()`` function to check how many species we have: 
```{r}
unique(iris$Species)
```
Now, we can use ``Species`` to give more insides into our data set. For example, we may be interested in computing the mean of ``Sepal.Lenght`` by species. For that, we can use the function ``tapply()`` as follows: 
```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
```
Or we can use ``tapply()`` in conjunction with ``summary()``.
```{r}
tapply(iris$Sepal.Length, iris$Species, summary)
```


We may also be interested in seeing if there is a relationship among the different columns of our data set. For instance, we can start by computing their correlation, a measure of how a pair of variables are linearly related. Recall that for two vectors $\mathbf{x} = (x_1, \dots, x_n)$ and $\mathbf{y} = (y_1, \dots, y_n)$ the sample covariance  is given by 
$$
\mbox{Cov}(\mathbf{x}, \mathbf{y}) = \frac{1}{n-1} \sum_{i = 1}^{n} \frac{(x_i - \bar{x})(y_i - \bar{y})}{s_{x}s_{y}} \,,
$$
where $s_{\mathbf{x}}$ and $s_{\mathbf{y}}$ are the sample standard deviations of $\mathbf{x}$ and $\mathbf{y}$, respectively. In R, the sample correlation is computed using the ``cor()`` function. For example, 
```{r}
cor(iris$Sepal.Length, iris$Sepal.Width)
```
If we have several (numeric) columns, we can compute all correlations at once:
```{r}
cor(iris[, -5])
```

:::{.remark}
The ``cor()`` computes, by default, Person's correlation coefficient described above. However, we can use the ``method`` argument to calculate Kendall's tau and Spearman's rho, two measures that assess how well the relationship between two variables can be described using a monotonic function.
:::


### Visual tools

We now present some R tools that can be employed to visualize data. More specifically, we will see how to create histograms, box plots, and scatter plots. 

#### Histograms {-}

To create a histogram, we can use the ``hist()`` function. For instance, to draw a histogram for ``Sepal.Length``, we type:
```{r}
hist(iris$Sepal.Length)
```

We can customize further our plot using the different argument of ``hist()`` (see ``?hist``):
```{r}
hist(iris$Sepal.Length,
  breaks = 15, # Number of cells in the histogram
  main = "Histogram of Sepal Lenght", # Main tittle
  xlab = "Sepal Lenght", # Text on the x-axis
  freq = FALSE, # Probability density
  xlim = c(4, 8), # Range for the x-axis
  col = "blue" # Color to fill the bars
)
```

R also allows to compute kernel density estimates via the ``density()`` function.  
```{r}
plot(density(iris$Sepal.Length))
```

We can combine our two previous plots using ``lines()`` as follows: 

```{r}
hist(iris$Sepal.Length,
  breaks = 15, # Number of cells in the histogram
  main = "Histogram of Sepal Lenght", # Main tittle
  xlab = "Sepal Lenght", # Text on the x-axis
  freq = FALSE, # Probability density
  xlim = c(4, 8), # Range for the x-axis
  col = "blue" # Color to fill the bars
)
lines(density(iris$Sepal.Length),
  lwd = 3, # Line width
  col = "red"
)
```


#### Box plots {-}
To produce box plots, R uses the ``boxplot()`` function. Let us try an example with the ``iris`` data set: 
```{r}
boxplot(iris$Sepal.Length)
```

We can create box plots for the different (numeric) columns of ``iris`` at the same time:

```{r}
boxplot(iris[, -5])
```


Let us now customize the above plot: 
```{r}
boxplot(iris[, -5],
  main = "Box plot - Iris data set",
  names = c("Sepal length", "Sepal width", "Petal length", "Petal width"), # Change names of x axis
  col = "green"
)
```


#### Scatter plots {-}
Scatter plots can help us to visualize the relationship between the different data columns. To create an scatter plot, we can use the ``plot()`` function. For instance, below we plot Sepal Length against Petal Length:

```{r}
plot(iris$Sepal.Length, iris$Petal.Length,
  panel.first = grid(8, 8), # Adds a 8x8 grid
  cex = 1.2, # Size of the dots
  col = "blue",
  xlab = "Sepal length",
  ylab = "Petal length",
  main = "Iris - Sepal length and petal length"
)
```

Finally, for data with several columns, we can plot them all at once:
```{r}
plot(iris[, -5],
  col = "blue",
)
```


## Probability distributions

R comes with several parametric probability distributions that can help us to describe our data. Every distribution in R has four functions, which can be called by using the *root* name of the distribution (e.g., ``exp``) preceded by one of the following letters: 

- ``d``: The probability density function (pdf).

- ``p``: The cumulative distribution function (cdf).

- ``q``: Quantile.

- ``r``: Random generator for the specified distribution. 


For instance, the exponential distribution has the four functions: ``dexp()``, ``pexp()``, ``qexp()`` and ``rexp()``.  

Table \@ref(tab:contdist) shows the continuous distributions,  and Table \@ref(tab:disdist) the discrete distributions available in R. Note, however, that these two lists are far from comprehensive. Fortunately, other distributions may be available in other packages. For example, an implementation for Pareto distribution can be found in the ``actuar`` R package.

``` {r, contdist, echo=FALSE}
d <- cbind(c("Beta", "Cauchy", "Chi-2", "Exponential", "Fisher F", "Gamma", "Logistic", "Lognormal", "Normal", "Student t", "Uniform", "Weibull"), c("beta", "cauchy", "chisq", "exp", "f", "gamma", "logis", "lnorm", "norm", "t", "unif", "weibull"))

knitr::kable(d, caption = "Continuous distributions.", booktabs = TRUE, col.names = c("Distribution", "Root name"))
```

``` {r, disdist, echo=FALSE}
d <- cbind(c("Binomial", "Geometric", "Hypergeometric", "Negative Binomial", "Poisson"), c("binom", "geom", "hyper", "nbinom", "pois"))

knitr::kable(d, caption = "Discrete distributions.", booktabs = TRUE, col.names = c("Distribution", "Root name"))
```

:::{.remark}
When using these R built-in functions, you must keep in mind that the parametric form of these distributions may differ from the ones you are familiar with. For instance, in R, the pdf of the Gamma distribution is implemented as (see help - ``?pgamma``):

$$
f(x) = \frac{1}{s^a \Gamma(a)} x^{a-1} e^{-x/s} \,, \quad x\ge 0\,, 
$$
where $a>0$ and $s>0$. In the other hand, some textbooks use

$$
f(x) = \frac{\lambda^\alpha}{ \Gamma(\alpha)} x^{\alpha-1} e^{-\lambda x} \,, \quad x\ge 0\,, 
$$
where $\lambda>0$ and $\alpha>0$. However, note that both representations are equivalent. Indeed, take $s = 1/\lambda$ and $a = \alpha$.
:::

We now look at some examples of how to use these functions, and we will use the normal distribution to do so. Recall that a random variable $X$ is said to be normal distributed with mean $\mu \in \mathbb{R}$ and standard deviation $\sigma>0$, if its density function is given by
$$
f(x) = \frac{1}{\sigma \sqrt{2 \pi }} \exp\left({- \frac{(x - \mu)^2}{2 \sigma^2}}\right) \,, \quad x\in \mathbb{R}\,. 
$$
We write $X \sim N(\mu, \sigma^2)$.

In R, this density can be evaluated using the ``dnorm()`` function. For instance,

```{r}
dnorm(1) # mu = 0, sigma = 1
dnorm(1, 1, 2) # mu = 1, sigma = 2
```


We will now plot this density function for different combinations of parameters. 
The usual way to plot a function in R is first generate a sequence of numbers, then evaluate the desired function at the generated sequence, and finally use ``plot()``.

```{r}
sq <- seq(-5, 5, by = 0.01)
plot(sq, dnorm(sq), type = "l", ylim = c(0, 0.85), ylab = "f(x)", xlab = "x")
lines(sq, dnorm(sq, 0, 0.5), type = "l", col = "red")
lines(sq, dnorm(sq, 0, 2), type = "l", col = "blue")
lines(sq, dnorm(sq, 1, 1), type = "l", col = "green")
lines(sq, dnorm(sq, -1, 1), type = "l", col = "orange")
legend("topright",
  leg = paste("mu =", c(0, 0, 0, 1, -1), ", sigma =", c(1, 0.5, 2, 1, 1)),
  lty = 1,
  col = c("black", "red", "blue", "green", "orange")
)
```

Next, we consider the distribution function of  $X \sim N(\mu, \sigma^2)$. Recall that this is given by

$$
F(x) = \mathbb{P} (X \leq x) = \int_{-\infty}^x \frac{1}{\sigma \sqrt{2 \pi }} \exp\left({- \frac{(y - \mu)^2}{2 \sigma^2}}\right) dy \,.
$$
To evaluate this function in R we can use ``pnorm()``. For example, if we want to find $\mathbb{P} (X \leq 1.8)$ for $X \sim N(0, 1)$, we simply type:
```{r}
pnorm(1.8)
```

Note that ``pnorm()`` has an optional argument ``lower.tail``. By default ``lower.tail`` takes the value ``TRUE``, indicating that ``pnorm()`` will compute $\mathbb{P} (X \leq x)$. However, we can compute $\mathbb{P} (X > x) = 1 - \mathbb{P} (X \leq x)$ by using ``lower.tail = FALSE``. For instance, 

```{r}
1 - pnorm(1.8)
pnorm(1.8, lower.tail = FALSE)
```

We can simulate random values following a $N(\mu, \sigma^2)$ distribution via the ``rnorm()`` function. For example, 
```{r, echo=FALSE}
set.seed(10)
```

```{r}
rnorm(10)
```

Note that every time you run ``rnorm()``, it will generate a different set of values: 

```{r}
rnorm(10)
```

However, sometimes it is convenient to generate the same random sequence at will. This can allow us, for example, to replicate a study. To do this, we need to give R an initial *seed* to generate the random numbers, which is done via the ``set.seed()`` function.  For example, type

```{r}
set.seed(1)
rnorm(10)
```

If we run the code again, we will get exactly the same sequence of values: 
```{r}
set.seed(1)
rnorm(10)
```

A visual way to evaluate if certain data follows an specific distribution, is to compare the histogram generated by the data and the density function of the distribution. Let us look at a simple example: First we generate a random sample following a $N(0, 1)$ distribution. 
```{r}
set.seed(1)
x <- rnorm(1000)
```

Now, if the data would really follow a $N(0, 1)$ distribution (which in this case it does by construction), the histogram of the data should be similar to the density of a $N(0, 1)$ distribution. The following plot shows exactly this: 
```{r}
hist(x, freq = F, main = "Histogram vs density")
sq <- seq(-5, 5, by = 0.01)
lines(sq, dnorm(sq), type = "l", col = "red")
```



Finally, we will look at function to compute quantiles of a distribution. Recall that for a random variable $X$ with distribution function $F$, its quantile function $F^{\leftarrow}$ is given by
$$
F^{\leftarrow}(p) = \inf \{x : p \leq F(x)  \} \,, \quad p \in [0,1] \,.
$$
If the distribution function is continuous and strictly monotonic, then $F^{\leftarrow}(p)$ satisfies that
$$
F(F^{\leftarrow}(p)) = p \,.
$$
In other words, it corresponds to the inverse of $F$. 

Now, consider $X \sim N(1, 2^2)$ and suppose that we want to find $x$ such that $P(X \leq x) = 0.95$. This can be done with the ``qnorm()`` function: 

```{r}
x <- qnorm(0.95, 1, 2)
x
pnorm(x, 1, 2) # Check
```

An alternative visual way of evaluating if our data comes from a certain distribution, is to plot the sample quantiles against the theoretical quantiles of the distribution to check. This is called, a QQ plot. If the data comes from that specific distribution, then we should observe that the points form (approximately) an identity line. Let us look at an example: 

```{r}
set.seed(1)
x <- rnorm(1000, 1, 2)
p <- seq(0.01, 0.99, by = 0.01)
the_quantile <- qnorm(p, 1, 2)
sam_quantile <- quantile(x, p)
plot(the_quantile, sam_quantile,
  main = "QQ plot",
  xlab = "Theoretical quantiles",
  ylab = "Sample quantiles"
)
abline(0, 1, col = "blue") # Identity
```


### Tranformations 

Several distributions are obtained via transformations. For instance, a lognormal distributed random variable can be obtained by exponentiation of a normal distributed random variable. Here, we will illustrate how to use the previous implementations to work with transformations of random variables. 

First let us recall some results. Let $Y$ be a continuous random variable with density and distributions functions $f_{Y}$ and $F_{Y}$, respectively. Now, consider a strictly increasing transformation $g(\cdot)$, and define $X = g(Y)$. Then, the distribution function $F_X$ of X is given by
$$
F_X(x) = \mathbb{P}(X \leq x) = \mathbb{P}(g(Y) \leq x) = \mathbb{P}(Y \leq g^{-1}(x)) = F_Y(g^{-1}(x)) \,.
$$
From the expression above, it follows that the density function $f_X$ of $X$ is given by
$$
f_X(x) =   f_Y(g^{-1}(x)) \frac{d}{dx} (g^{-1}(x)) \,.
$$

Now, let us consider an explicit example: the lognormal distribution. Recall that a lognormal distributed random variable with parameters $\mu$ and $\sigma^2$ is obtained via the transformation

$$
X = \exp(Y) \,,
$$
where $Y \sim N(\mu, \sigma^2)$. We write $X \sim LN(\mu, \sigma^2)$. Although already implemented in R under the root name ``lnorm``, we can use this distribution to illustrate how to work with transformations, and at the same time, verify our computations. Note that in this case $g(y) = \exp(y)$, $y \in \mathbb{R}$, and $g^{-1}(x) = \log(x)$, $x>0$. 

We can now compute the distribution function $F_X(x)$ of $X \sim LN(\mu, \sigma^2)$ at $x$, by using ``pnorm()`` evaluated at $g^{-1}(x) = \log(x)$. For example, for $X \sim LN(1, 2^2)$, $\mathbb{P}(X \leq 3)$ can be computed as follows:
```{r}
user_plnorm <- function(x, mu, sigma) {
  pnorm(log(x), mu, sigma)
}
user_plnorm(3, 1, 2)
plnorm(3, 1, 2) # Check
```
Now for the density evaluation, we need $\frac{d}{dx}(g^{-1}(x)) = 1/x$. Thus, the density evaluation at $x = 3$ for $X \sim LN(1, 2^2)$ can be computed as: 
```{r}
user_dlnorm <- function(x, mu, sigma) {
  dnorm(log(x), mu, sigma) / x
}
user_dlnorm(3, 1, 2)
dlnorm(3, 1, 2) # Check
```


In this case, since our transformation is strictly monotonic, we can compute the quantiles easily. 
```{r}
user_qlnorm <- function(p, mu, sigma) {
  exp(qnorm(p, mu, sigma))
}
user_qlnorm(0.95, 1, 2)
qlnorm(0.95, 1, 2) # Check
```



Simulation can also be performed using the relationship $X = \exp(Y)$
```{r}
user_rlnorm <- function(n, mu, sigma) {
  exp(rnorm(n, mu, sigma))
}
set.seed(1)
user_rlnorm(10, 1, 2)
```
In this case, we can check that our simulation is correct by using ``rlnorm()``
```{r}
set.seed(1)
rlnorm(10, 1, 2)
```




### Law of large numbers and central limit theorem

We now recall two important results in probability theory and illustrate them using R, namely the *Law of Large Numbers* and the *Central Limit Theorem*. We start with the Law of Large Numbers:

:::{.theorem}
Let $X_1,X_2,\dots$ be a sequence of i.i.d. random variables such that $\mathbb{E}[X_1]=\mu<\infty$. Then, for any $\epsilon >0$ 
$$
\underset{n\rightarrow \infty}{\lim}P\{|\bar{X}_n-\mu|\geq \epsilon\}=0 \,,
$$
where $\bar{X}_n = n^{-1}\sum_{i = 1}^n X_i$ denotes the sample mean.
In other words, as the sample size $n$ increases, the probability that the sample mean $\bar{X}_n$ is different from the population mean $\mu$ converges to zero. 

:::

We now illustrate this result via a simulation:

```{r}
n <- 1000
plot(cumsum(rnorm(n)) / 1:n,
  xlab = "Sample size",
  ylab = "Sample mean",
  col = "darkgray"
)
abline(0, 0, col = "blue")
```

In the figure above, we can observe that as the sample size increases, the sample mean gets closer to the population mean (zero). 
Next, we recall the Central Limit Theorem:

::: {.theorem}
Let $X_1,X_2,\ldots$ be a sequence of i.i.d. random variables with a finite expected value $\mathbb{E}[X_1]=\mu<\infty$ and variance $\mbox{Var}[X_1] = \sigma^2 <\infty$. Now, let $Z_n$ be the standarized mean given by
$$
Z_n:=\frac{\bar{X}_n-\mu}{\sigma/\sqrt{n}}.
$$
Then, for $n$ sufficiently large,
$$
 Z_n\simeq N(0,1).
$$
Or equivalently, 
$$
\bar{X}_n\simeq N(\mu,\sigma^2/n)\,,\quad 
$$
In other words,	for sufficiently large $n$, the sample mean $\bar{X}_n$ is close to be normal distributed with mean $\mu$ and variance $\sigma^2/n$.

:::

Let us exemplify this result via a simulation study. The idea is as follows: consider a sample of size $n$ from an exponential distributed random variable with mean $1/\lambda$. If we were to observe several samples of size $n$ from this distribution, lets say $p$ samples, and compute $Z_n^{(j)}$ for each sample $j = 1,\dots, p$, then for $n$ large enough $Z_n^{(j)}$, $j = 1,\dots, p$, should be approximately a sample from a standard normal distributed random variable. We could then, for example, check if $Z_n^{(j)}$, $j = 1,\dots, p$ is truly normal distributed via a QQ plot. Let us try this, with $n = 10$, $p = 500$ and $\lambda = 0.5$:

```{r}
n <- 10 # Sample size
p <- 500 # Replications of experiment
lambda <- 0.5
sim_exp <- matrix(rexp(n * p, lambda), p, n) # Simulations in matrix form
samp_mean <- apply(sim_exp, 1, mean) # Sample mean for each replication
mu <- 1 / lambda # Population mean
sigma2 <- 1 / lambda^2 # Population variance
std_mean <- (samp_mean - mu) / (sqrt(sigma2 / n)) # Standardized mean (Z_n)
qqnorm(std_mean, main = "Normal QQ plot - n = 10") # QQ plot with standard normal
abline(0, 1, col = "blue")
```

In the figure above, we observe that $Z_n$ is not standard normal distributed (yet). However, if we now consider $n = 1000$, the conclusion changes (see below).

```{r, echo=FALSE}
n <- 1000 # Sample size
p <- 500 # Replications of experiment
lambda <- 0.5
sim_exp <- matrix(rexp(n * p, lambda), p, n) # Simulations in matrix form
samp_mean <- apply(sim_exp, 1, mean) # Sample mean for each replication
mu <- 1 / lambda # Population mean
sigma2 <- 1 / lambda^2 # Population variance
std_mean <- (samp_mean - mu) / (sqrt(sigma2 / n)) # Standardized mean (Z_n)
qqnorm(std_mean, main = "Normal QQ plot - n = 1000") # QQ plot with standard normal
abline(0, 1, col = "blue")
```


## Parametric inference {#parinf}

In the last section, we presented some distributions in R that can be used to model our data. However,  so far, we have not covered how to fit these models to given data, which is an essential part for their application in insurance and finance. Therefore, this section aims to present some estimation methods available. 


### Maximum likelihood estimation 

Let $\boldsymbol{X} = (X_1, \dots, X_n)$ be a random sample such that $X_i$ are i.i.d. random variables with common distribution function $F(\cdot; \boldsymbol{\theta})$, where $\boldsymbol{\theta} = (\theta_1,\dots,\theta_d) \in \boldsymbol{\Theta} \subset \mathbb{R}^d$. Here, $\boldsymbol{\Theta}$ is known as the parameter space. Given an observed data sample $\mathbf{x} = (x_1, \dots, x_n)$ from $\boldsymbol{X}$, the *likelihood function* $L$ is defined as the joint density (as a function of $\boldsymbol{\theta}$) evaluated at $\mathbf{x}$, that is, 

$$
L(\boldsymbol{\theta}; \mathbf{x}) = \prod_{i = 1}^{n} f(x_i; \boldsymbol{\theta}) = f_{\boldsymbol{X}} (\mathbf{x}; \boldsymbol{\theta}) \,.
$$
The *maximum likelihood estimator* (MLE) is defined as the value of $\boldsymbol{\theta}$ that maximizes the likelihood function, that is, 

$$
\hat{\boldsymbol{\theta}} = \mathrm{arg\,max}_{\boldsymbol{\theta} \in \boldsymbol{\Theta}} L(\boldsymbol{\theta}; \mathbf{x}) = \mathrm{arg\,max}_{\boldsymbol{\theta} \in \boldsymbol{\Theta}} \prod_{i = 1}^{n} f(x_i; \boldsymbol{\theta}) \,.
$$

In practice, it is often more convenient to work with the *log-likelihood function* $l$, which is obtained by taking $\log$ of the likelihood function, i.e., 
$$
l(\mathbf{x}, \boldsymbol{\theta}) = \log(L(\boldsymbol{\theta}; \mathbf{x})) = \sum_{i = 1}^{n} \log\left(f(x_i; \boldsymbol{\theta}) \right) \,.
$$
Since $\log$ is an increasing continuous function, maximizing the log-likelihood is equivalent to maximizing the likelihood. In other words, 
$$
\hat{\boldsymbol{\theta}} = \mathrm{arg\,max}_{\boldsymbol{\theta} \in \boldsymbol{\Theta}} l(\boldsymbol{\theta}; \mathbf{x}) = \mathrm{arg\,max}_{\boldsymbol{\theta} \in \boldsymbol{\Theta}} \sum_{i = 1}^{n} \log(f(x_i; \boldsymbol{\theta})) \,.
$$
Lets us now look at some examples on how to program log-likelihood functions. First, we simulate a sample from an exponential distributed random variable:
```{r}
set.seed(1)
lambda <- 1.5
x_exp <- rexp(1000, lambda)
```

The following implementation, computes the log-likelihood, assuming exponential distributed observations,  for different parameters:
```{r}
loglik_exp <- function(x, lambda) {
  sum(log(dexp(x, lambda)))
}
```

Given that our data, originally comes from an exponential distribution with $\lambda = 1.5$, we expect that values close to $1.5$ return bigger values.
```{r}
loglik_exp(x_exp, 0.5)
loglik_exp(x_exp, 1)
```

Moreover, we can plot the log-likelihood as a function of $\lambda$. To do so, we need to modify our log-likelihood implementation above to work with vector inputs for the parameter. 
```{r}
loglik_exp <- function(x, lambda) {
  ll <- rep(0, length(lambda))
  for (i in 1:length(lambda)) {
    ll[i] <- sum(log(dexp(x, lambda[i])))
  }
  ll
}
```
Now, we can generate our plot. 
```{r}
lambda_sq <- seq(0.01, 4, by = 0.01)
plot(lambda_sq, loglik_exp(x_exp, lambda_sq),
  main = "Log-likelihood",
  ylab = "l(lambda; x)",
  xlab = "Lambda",
  type = "l"
)
```

The figure above shows that the log-likelihood is maximized around the original value $\lambda = 1.5$ (as expected). 


Let us now consider a second example, where we require to estimate two parameters, namely a normal distribution. First, we simulate a sample:
```{r}
set.seed(1)
mu <- 1
sigma <- 2
x_norm <- rnorm(1000, mu, sigma)
```
Next, we implement the log-likelihood:
```{r}
loglik_norm <- function(x, mu, sigma) {
  sum(log(dnorm(x, mu, sigma)))
}
```
Let us test our function:
```{r}
loglik_norm(x_norm, 0.5, 1)
loglik_norm(x_norm, 0.5, 1.5)
loglik_norm(x_norm, 0.8, 1)
loglik_norm(x_norm, 0.8, 1.5)
loglik_norm(x_norm, 2.5, 2.5)
```
Again, we can see that we obtain larger values close to the real parameters $\mu = 1$ and $\sigma = 2$. We now modify the function for plotting purposes.
```{r}
loglik_norm <- function(x, mu, sigma) {
  ll <- rep(0, length(mu))
  for (i in 1:length(mu)) {
    ll[i] <- sum(log(dnorm(x, mu[i], sigma[i])))
  }
  ll
}
```
Now, we can create a surface plot. For that purpose, we need to create a grid of evaluation points and evaluate our log-likelihood at all those point. This can be done as follows:
```{r, fig.align = 'center'}
mu_sq <- seq(0.75, 1.5, by = 0.05)
sigma_sq <- seq(0.75, 2.5, by = 0.05)
ll_eval <- outer(mu_sq, sigma_sq, loglik_norm, x = x_norm) # Evaluates the Log-Lik at all points
persp(mu_sq, sigma_sq, ll_eval,
  phi = 20,
  theta = 45,
  ticktype = "detailed",
  xlab = "mu",
  ylab = "sigma",
  zlab = "",
  main = "Log-likelihood"
)
```

#### Maximization {-}

Our next step is to maximize the log-likelihood with respect to the parameters. In R, this can be done by using ``optim()``, which is a function that performs minimization. Remember that maximizing a function is equivalent to minimizing the negative of that function. Thus, we need to work with the negative log-likelihood in order to use ``optim()``. Let us exemplify the above with the exponential distribution.

Consider the previous exponential sample (``x_exp``). Next, we implement the negative log-likelihood:
```{r}
nloglik_exp <- function(x, lambda) {
  -sum(log(dexp(x, lambda)))
}
```
With this implementation at hand, we can find the MLE of $\lambda$ using ``optim()``.
```{r}
mle_exp <- optim(
  par = 1, # Initial value for the parameter to be optimized over
  fn = nloglik_exp, # Function to be minimized
  x = x_exp # Further parameters
)
mle_exp$par # MLE - note that it is close to the original parameter
mle_exp$value # Negative log-likelihood
```
In this particular case, we can verify our result using that the MLE for the exponential distribution has an explicit solution given by 
$$
\hat{\lambda} = 1 / \bar{x}_n \,.
$$
With our simulated sample, we then obtain
```{r}
lambda_hat <- 1 / mean(x_exp)
lambda_hat
```
which is very close to our solution with ``optim()``.


Let us now try the same procedure with our normal distributed sample. In this case, we require to estimate two parameters. The first step is to implement the negative log-likelihood. 
```{r}
nloglik_norm <- function(x, par) {
  -sum(log(dnorm(x, par[1], par[2])))
}
```
Note that in the above function, we passed both parameters as a single argument. We need to do it in this way in order to be able to use ``optim()``.

```{r}
mle_norm <- optim(
  par = c(0.5, 1.5), # Initial values for the parameters to be optimized over
  fn = nloglik_norm,
  x = x_norm
)
mle_norm$par # MLE - note that it is close to the original parameter
mle_norm$value # Negative log-likelihood
```

Again, we can verify our results using that for the normal distribution, the MLE's of $\mu$ and $\sigma$ are explicit and given by
$$
\hat{\mu} = \bar{x}_n \,, \quad \hat{\sigma} = \left( n^{-1} \sum_{i = 1}^n(x_i - \bar{x}_n)^2 \right)^{1/2}
$$
Thus, for our simulated sample, we obtain 
```{r}
mu_hat <- mean(x_norm)
sigma_hat <- sqrt(sum((x_norm - mean(x_norm))^2) / length(x_norm))
mu_hat
sigma_hat
```
Note that our results using ``optim()`` are very close to the ones obtained via the closed-form formulas. 


#### Properties of the MLE {-}

We now review some properties of maximum likelihood estimators. For illustration purposes, we state the results when the parameter space is a subset of $\mathbb{R}$. However, the results can be extended to higher dimensions. 

##### Consistency {-}
:::{.theorem}
Let $X_1,X_2,\dots$ be a sequence of i.i.d. random variables with common density function $f(\cdot;\theta)$. Then, under mild conditions, for any $\epsilon >0$ 
$$
\underset{n\rightarrow \infty}{\lim}P\{|\hat{\theta}_n-\theta|\geq \epsilon\}=0 \,,
$$
where $\hat{\theta}_n$ is the maximum likelihood estimator of $\theta$ based on a sample of size $n$.
In other words, as the sample size $n$ increases, the probability that the MLE $\hat{\theta}_n$ is different from the true parameter $\theta$ converges to zero. 
:::

Let us now illustrate this with a simulation.
```{r}
set.seed(1)
x_10 <- rexp(10, rate = lambda)
x_100 <- rexp(100, rate = lambda)
x_1000 <- rexp(1000, rate = lambda)
mle_exp_10 <- suppressWarnings(optim(par = 1, fn = nloglik_exp, x = x_10)$par)
mle_exp_100 <- suppressWarnings(optim(par = 1, fn = nloglik_exp, x = x_100)$par)
mle_exp_1000 <- suppressWarnings(optim(par = 1, fn = nloglik_exp, x = x_1000)$par)
mle_exp_10
mle_exp_100
mle_exp_1000
```
We can see that as the sample size increases, the MLE gets closer to the true parameter. 

##### Efficiency {-}


::: {.theorem #effmle}
Let $X_1,X_2,\ldots$ be a sequence of i.i.d. random variables with common density function $f(\cdot;\theta)$. Now, let $\hat{\theta}_n$ be the maximum likelihood estimator of $\theta$ based on a sample of size $n$. Then, under mild conditions, for $n$ sufficiently large,

\begin{equation} 
 \sqrt{n}(\hat{\theta}_n - \theta)\simeq N(0,(\mathcal{I}(\theta))^{-1})\,,
  (\#eq:anmle)
\end{equation} 

where  $\mathcal{I}(\theta)$ is the Fisher information
$$
\mathcal{I}(\theta) = \mathbb{E}_\theta\left[ -\frac{d^2}{d\theta^2} \log(f(X;\theta)) \right] \,.
$$
:::

We now give an equivalent representation for  \@ref(eq:anmle). First, we define the information of the sample $\mathcal{I}_n(\theta)$ as 
$$
\mathcal{I}_n(\theta) = n\mathcal{I}(\theta) =  \mathbb{E}_\theta\left[ -\frac{d^2}{d\theta^2} \log(L(\theta;\mathbf{X})) \right] \,.
$$
Next, we define the *standard error* $se$ as 
$$
se = \sqrt{1 / \mathcal{I}_n(\theta)}
$$
Then, \@ref(eq:anmle) can be rewritten as

\begin{equation} 
 \frac{(\hat{\theta}_n - \theta)}{se}\simeq N(0,1)\,,
  (\#eq:anmle2)
\end{equation}

In practice, $\mathcal{I}_n(\theta)$ can be approximated by the observed information $\hat{\mathcal{I}}_n(\hat{\theta}_n)$ given by 

$$
\hat{\mathcal{I}}_n(\hat{\theta}_n) =  -\frac{d^2}{d\theta^2} \log(L(\theta;\mathbf{x}))\mid_{\theta = \hat{\theta}_n}  \,.
$$
Consequently, the standard error can be approximated as $\hat{se} = \sqrt{1 / \mathcal{I}_n(\hat{\theta}_n)}$. 

In R, we can compute $\hat{\mathcal{I}}_n(\hat{\theta}_n)$ by using the argument ``hessian`` in ``optim()``. For example,

```{r}
mle_exp <- optim(
  par = 1, # Initial value for the parameter to be optimized over
  fn = nloglik_exp, # Function to be minimized
  hessian = TRUE, # Computes the Hessian
  x = x_exp # Further parameters
)
obs_inf_exp <- mle_exp$hessian
obs_inf_exp
```
Then, we can approximate the standard error as well.
```{r}
se_exp <- sqrt(1 / obs_inf_exp)
se_exp
```


Now, we can use the asymptotic normality of the MLE to compute confidence intervals. More specifically, an $(1 -\alpha)$ confidence interval for $\theta$ is given by 
$$
\hat{\theta}_n \pm q_{(1 -\alpha/2)}\, \hat{se} \,,
$$

where $q_{(1 -\alpha/2)}$ is the $(1 -\alpha/2)$-quantile of the standard normal distribution. 



##### Further properties of the MLE {-}
The following theorem provides further properties of the MLE

:::{.theorem}
$\,$

a) Let $\hat{\boldsymbol{\theta}}$ be the MLE of $\boldsymbol{\theta}$. Then, given a function $g : \mathbb{R}^d \to \mathbb{R}$,  the MLE of $g(\boldsymbol{\theta})$ is  $g(\hat{\boldsymbol{\theta}})$. This is called the invariance property of the MLE. 
b) If $\mathbf{Y} = \mathbf{h} (\mathbf{X})$, where $\mathbf{h}$ is invertible in the domain of $\mathbf{X}$, then the MLE based on $\mathbf{Y}$ is the same as the MLE based on $\mathbf{X}$. 
:::


#### R packages for MLE {-}

There are several R packages that include functions to perform maximum likelihood estimation. For instance, ``EstimationTools``, ``fitdistrplus``, ``MASS``, among others. Here, we illustrate the use of the ``fitdistrplus`` package.
```{r}
library(fitdistrplus)
```

The ``fitdistrplus`` package comes with the ``fitdist()`` function to perform MLE (and other estimation methods). The distribution is specified with the argument ``distr``, but the density (``d``) and distribution (``p``) functions must be available for the given distribution (see help for details ``?fitdist``). Note that the argument ``start`` can be omitted for some distributions (the list can be found in the help), otherwise it has to be specified.  Let us give a couple of examples.

First, MLE for our exponential sample:
```{r}
fit_exp <- fitdist(x_exp, distr = "exp")
summary(fit_exp)
```

Secondly, MLE for our normal sample:
```{r}
fit_norm <- fitdist(x_norm, distr = "norm")
summary(fit_norm)
```

Note that we obtain (approximately) the same values as our implementations above.

:::{.remark}
Another useful implementation in ``fitdistrplus`` is that it can generate plots to evaluate the quality of the fit by simply using the ``plot()`` function. For example, 
```{r}
plot(fit_exp)
```
:::


Finally, we provide an example with a user-defined distribution, namely, the Gumbel distribution with location parameter $\mu \in \mathbb{R}$ and scale parameter $\beta > 0$. Recall that the density and distribution functions of this model are given by 
$$
f(x) = \exp\left(-\exp\left(-\frac{(x - \mu)}{\beta}\right)\right)\exp\left(-\frac{(x - \mu)}{\beta}\right) \frac{1}{\beta} \,, \quad x\in\mathbb{R} \,,
$$
$$
F(x) = \exp\left(-\exp\left(-\frac{(x - \mu)}{\beta}\right)\right)\,, \quad x\in\mathbb{R} \,.
$$
An implementation of these functions is the following:

```{r}
dgumbel <- function(x, mu, beta) {
  exp((mu - x) / beta) * exp(-exp((mu - x) / beta)) / beta
}
dgumbel(1, 1, 2)
pgumbel <- function(q, mu, beta) {
  exp(-exp((mu - q) / beta))
}
pgumbel(1, 1, 2)
```

With these implementations at hand, can now call ``fitdist()`` to perform MLE. For example, we consider the ``groundbeef`` data set in the ``fitdistrplus`` package
```{r}
data(groundbeef)
```
and we fit a Gumbel distribution to the serving sizes
```{r}
fit_gum <- fitdist(groundbeef$serving,
  distr = "gumbel",
  start = list(mu = 1, beta = 2)
)
summary(fit_gum)
```

```{r, echo=F}
rm(list = ls())
```


### Adequacy of the fit
We now focus on measuring the adequacy of a fit either via graphical methods or numerical methods. To illustrate the methods, we consider the Danish fire insurance data set (``danishuni``) available in ``fitdistrplus``. More specifically, we will consider the losses above 1 million danish krone and subtract 1 million to all data points to bring the data to the origin, that is, 
```{r}
data(danishuni)
danish_loss <- danishuni$Loss[danishuni$Loss > 1] - 1
summary(danish_loss)
```

We will consider the following potential models for the transformed data: 

a) Gamma - ``gamma``
b) Pareto - ``pareto`` (available in the ``actuar`` R package)
c) Burr - ``burr`` (available in the ``actuar`` R package)
d) Mixture of Gamma and Pareto - To be implemented 

The idea is to select the model that "best" described the data from the above list. 

 First, we need to load the ``actuar`` package to get access to the implementations of the Pareto and Burr distributions.
```{r}
library(actuar)
```
We can now perform MLE for the Gamma, Pareto and Burr distributions
```{r}
fit_gamma <- fitdist(danish_loss,
  distr = "gamma"
)
fit_burr <- fitdist(danish_loss,
  distr = "burr",
  start = list(shape1 = 1, shape2 = 1, scale = 1)
)
fit_pareto <- fitdist(danish_loss,
  distr = "pareto",
  start = list(shape = 1, scale = 1)
)
```
To perform MLE of the mixture of Gamma and Pareto, we need to program its density and distribution functions. This is easily done using the corresponding functions for the ``gamma`` and ``pareto`` distributions: 
```{r}
dmgp <- function(x, shapeg, rateg, shapep, scalep, prob) {
  prob * dgamma(x, shapeg, rateg) + (1 - prob) * dpareto(x, shapep, scalep)
}
pmgp <- function(q, shapeg, rateg, shapep, scalep, prob) {
  prob * pgamma(q, shapeg, rateg) + (1 - prob) * ppareto(q, shapep, scalep)
}
dmgp(1, 1, 1, 1, 1, 0.5)
pmgp(1, 1, 1, 1, 1, 0.5)
```
We can now perform MLE for the above model: 
```{r}
fit_mgp <- fitdist(danish_loss,
  distr = "mgp",
  start = list(shapeg = 1, rateg = 1, shapep = 1, scalep = 1, prob = 0.5),
  lower = 0 # Avoid negative values
)
```

:::{.remark}
Although numerical maximization of mixture models can be done directly, aka by "brute force," there are more efficient ways to perform MLE of these models. For example, by using the expectation-maximization (EM) algorithm.
:::

Now, let us look at the results of our fits: 
```{r}
summary(fit_gamma)
summary(fit_pareto)
summary(fit_burr)
summary(fit_mgp)
```

The first number we can look at to select a model is the likelihood. Remember that we aim to maximize the likelihood; hence we would prefer a model with the highest likelihood possible. In our case, the mixture model has the highest likelihood. However, we also need to take into account the complexity of a model. Let us explain the last point in detail: In general, statistical models with more parameters allow for more flexibility, and thus, we can expect improvements in the fits as the number of parameters increases. However, having too many parameters can result in a fit that only describes the data at hand and fails to fit additional data or predict future observations. This is known as overfitting. On the other hand, underfitting occurs when a model cannot adequately capture the underlying structure of the data. Information criteria such as the Akaike information criterion (AIC) and the Bayesian information criterion (BIC), deal with the problem of overfitting by introducing a penalty term for the number of parameters in the model. More specifically, if we let $\hat {L}$ be the maximum value of the likelihood function, $d$ the number of parameters in our model, and $n$ the sample size, the AIC and BIC are computed as

$$ \mathrm {AIC} \,=\,2d-2\ln({\hat {L}})$$
$$ \mathrm {BIC} \,=\,d\log(n)-2\ln({\hat {L}})$$
Given a set of candidate models for our data, the preferred model would be the one with the minimum AIC (or BIC) value. In our current example, these number indicate that the mixture model is still preferred. 


We can also use visual tools to assess the quality of the fit. Let us start by plotting the histogram of the data 
```{r}
hist(danish_loss, freq = F, breaks = 30)
```

Given that our data has losses that are quite large, it may be more convenient to plot the logarithm of the data
```{r}
hist(log(danish_loss), freq = F, breaks = 30)
```

We now compare the histogram with the fitted distributions. Note, however, that we need to adapt our density functions accordingly to compare with the logarithm of the data by using the change of variable theorem. 

```{r}
sq <- seq(-7, 7, by = 0.01)
hist(log(danish_loss), freq = F, breaks = 30)
lines(sq, dgamma(exp(sq), fit_gamma$estimate[1], fit_gamma$estimate[2]) * exp(sq), col = "red")
lines(sq, dpareto(exp(sq), fit_pareto$estimate[1], fit_pareto$estimate[2]) * exp(sq), col = "blue")
lines(sq, dburr(exp(sq), fit_burr$estimate[1], fit_burr$estimate[2]) * exp(sq), col = "green")
lines(sq, dmgp(exp(sq), fit_mgp$estimate[1], fit_mgp$estimate[2], fit_mgp$estimate[3], fit_mgp$estimate[4], fit_mgp$estimate[5]) * exp(sq))
```
We observe that the density mixture distribution is closer to the histogram. Finally, we can also create some QQ plots to evaluate the fit. Notice that to create the QQ plot of the mixture distribution, we need to implement a function to compute its quantiles. An implementation is the following:
```{r}
qmgp <- function(p, shapeg, rateg, shapep, scalep, prob) {
  L2 <- function(q, p) {
    (p - pmgp(q, shapeg, rateg, shapep, scalep, prob))^2
  }
  sapply(p, function(p) optimize(L2, c(0, 10^3), p = p)$minimum)
}
qmgp(0.5, 1, 1, 1, 1, 0.5)
```

We can now create our QQ plots:
```{r}
p <- seq(0.001, 0.999, by = 0.001)
danish_quant <- quantile(danish_loss, p)

par(mfrow = c(2, 2))
plot(qgamma(p, fit_gamma$estimate[1], fit_gamma$estimate[2]),
  danish_quant,
  main = "QQ plot - Gamma",
  xlab = "Quantiles fitted Gamma",
  ylab = "Empirical quantiles"
)
abline(0, 1, col = "blue")

plot(qpareto(p, fit_pareto$estimate[1], fit_pareto$estimate[2]),
  danish_quant,
  main = "QQ plot - Pareto",
  xlab = "Quantiles fitted Pareto",
  ylab = "Empirical quantiles"
)
abline(0, 1, col = "blue")

plot(qburr(p, fit_burr$estimate[1], fit_burr$estimate[2]),
  danish_quant,
  main = "QQ plot - Burr",
  xlab = "Quantiles fitted Burr",
  ylab = "Empirical quantiles"
)
abline(0, 1, col = "blue")

plot(qmgp(p, fit_mgp$estimate[1], fit_mgp$estimate[2], fit_mgp$estimate[3], fit_mgp$estimate[4], fit_mgp$estimate[5]),
  danish_quant,
  main = "QQ plot - Gamma-Pareto mixture",
  xlab = "Quantiles fitted Gamma-Pareto mixture",
  ylab = "Empirical quantiles"
)
abline(0, 1, col = "blue")
```


One of the difficulties of this data set is the large values, which make it challenging to find a model that adequately models the body and tail simultaneously. However, the problem can be split. On the one hand, one can find a model for the body of the distribution, and on the other, a model for the tail of the distribution by, for example, using Extreme value theory statistics. In fact, this data set has been analyzed extensively using Extreme value theory techniques.


### Other estimation methods
Although MLE is the most common estimation method, we quickly review some alternatives. It is worth remarking that the following methods are implemented in the ``fitdistrplus`` R package. Thus, we will illustrate them using that package.

##### Method of moments {-}
Let $X_1, \dots, X_n$ be i.i.d. random variables with common distribution function $F(\cdot; \boldsymbol{\theta})$, where $\boldsymbol{\theta} = (\theta_1,\dots,\theta_d) \in \boldsymbol{\Theta} \subset \mathbb{R}^d$. Assuming that the moments up to order $d$ of $X \sim F(\cdot; \boldsymbol{\theta})$ exist, we set
$$
\mu_k(\boldsymbol{\theta}) = \mathbb{E}\left[ X^k\right] = \mu_k \,, \quad k = 1,\dots, d\,,
$$
In particular, this means that given $\mu_1, \dots, \mu_d$ we have $d$ nonlinear equations with $d$ unknown $\theta$'s. Therefore, we can find $\boldsymbol{\theta}$ from the raw moments of $X$, assuming that a unique solution exists, which is often the case. In practice, we can use the empirical raw moments $\hat{\mu}_k$ given by 
$$
\hat{\mu}_k = \frac{1}{n} \sum_{i = 1}^n x_i^k\,, \quad k = 1,\dots, d\,,
$$
Then, the moment matching estimator $\hat{\boldsymbol{\theta}}$ of $\boldsymbol{\theta}$ is the solution to
$$
\mu_k({\boldsymbol{\theta}}) =  \hat{\mu_k} \,, \quad k = 1,\dots, d\,,
$$
Sometimes it is more convenient to base the estimation on central moments. Let
$$
m_1(\boldsymbol{\theta}) = \mathbb{E}\left[ X\right] \quad \mbox{and} \quad m_k(\boldsymbol{\theta}) = \mathbb{E}\left[ (X - \mathbb{E}\left[ X\right])^k\right] \,,k = 2,\dots, d
$$
and 
$$
\hat{m}_1 = \bar{x}_n \quad \mbox{and} \quad \hat{m}_k =  \frac{1}{n} \sum_{i = 1}^n (x_i - \bar{x}_n )^k\,, \quad k = 2,\dots, d\,,
$$
Then we can solve for $\boldsymbol{\theta}$ in
$$
m_k({\boldsymbol{\theta}}) =  \hat{m_k} \,, \quad k = 1,\dots, d\,,
$$
to find the moment matching estimator $\hat{\boldsymbol{\theta}}$.

In general, there are no closed-form formulas for the moment matching estimators, and numerical methods must be employed. In R, we can use, for example, the ``fitdist()`` function in the ``fitdistrplus`` R package to perform moment matching estimation by simply adjusting the argument ``method``. For example, for our exponential distributed sample
```{r}
set.seed(1)
lambda <- 1.5
x_exp <- rexp(1000, lambda)
```
we compute the moment matching estimator
```{r}
fit_exp_mme <- fitdist(x_exp, distr = "exp", method = "mme")
summary(fit_exp_mme)
```

:::{.remark}
You may have noticed that the maximum likelihood and the moment matching estimators coincide for our exponential distributed sample. This is a particular property of the exponential distribution. However, in general, this is not the case for other distributions.
:::


##### Quantile matching {-}

Quantile matching estimation follows a similar idea to the moment matching estimation. The main difference is that we now match the empirical quantiles of the and the theoretical quantiles of the parametric distribution to be fitted. More specifically, we have the following equations

$$
F^{\leftarrow}(p_k;\boldsymbol{\theta}) = Q_{n, p_k}\,, \quad k = 1, \dots, d \,,
$$
where $Q_{n, p_k}$ are the empirical quantiles for specified probabilities $p_k \in [0,1]$, $k = 1,\dots,d$. The solution  $\hat{\boldsymbol{\theta}}$ for the above equations is the quantile matching estimator and is (typically) computed using numerical optimization. Note that the probabilities $p_k \in [0,1]$, $k = 1,\dots,d$ must be pre-selected and, thus, the resulting estimator would depend on this selection. We can use the ``fitdist()`` function in R to perform this estimation method. We simply change the input of the argument ``method`` to "qme". Note that in this case, we need to specify the probabilities via the argument ``probs``. Let us give a specific example

```{r}
fit_exp_qme <- fitdist(x_exp, distr = "exp", method = "qme", probs = 0.5)
summary(fit_exp_qme)
```

##### Maximum Goodness-of-fit {-}

The last method that we present here is the Goodness-of-fit estimation method, also known as minimum distance estimation. The idea of this method is to find a parameter $\boldsymbol{\theta}$ that minimize the "distance" between the empirical distribution function $F_{n}(x) = n^{-1}\sum_{i = 1}^n \mathbf{1}(x_i \leq x)$ and the parametric distribution  $F(x; \boldsymbol{\theta})$. The distance can be measure, for example, using the Cramer–von Mises distance defined as
$$
D(\boldsymbol{\theta}) = \int_{-\infty}^{\infty} (F_{n}(x) - F(x; \boldsymbol{\theta}))^2 dx
$$
which in practice can be estimated as 

$$
\hat{D}(\boldsymbol{\theta}) = \frac{1}{12 n} + \sum_{i = 1}^{n} \left(F(x; \boldsymbol{\theta}) - \frac{2i - 1}{2n} \right)^2 dx
$$
Thus, maximum goodness-of-fit estimation translates to finding $\boldsymbol{\theta}$ that minimizes $\hat{D}(\boldsymbol{\theta})$. In R, we can access this method by changing the argument ``method`` to "mge" in the ``fitdist()`` function. Let us finish this section by trying this method in our exponential sample

```{r}
fit_exp_mge <- fitdist(x_exp, distr = "exp", method = "mge")
summary(fit_exp_mge)
```
Note that the Cramer-von Mises distance is the default option, but it can be changed to other distances (see help for details).


## Multivariate distributions


Let $\mathbf{X} = (X_1,\dots, X_p)$ be a random vector with joint distribution function $F_{\mathbf{X}}$ and marginal distribution functions $F_j$ (i.e.,  $X_j \sim F_j$), $j = 1,\dots ,p$. In this section we review some parametric models for $F_{\mathbf{X}}$. First, we review one the most well know multivariate model, namely the multivariate normal distribution and then we present a more general multivariate modeling approach using copulas. 


### Multivariate normal distribution

Recall that we say that a $p$-dimensional random vector $\mathbf{X} = (X_1,\dots, X_p)$ is said to be multivariate normal distributed with mean vector $\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$ if its joint density function $f_{\mathbf{X}}$ is given by 

$$
f_{\mathbf{X}}(x_1,\dots, x_p) = \frac{1}{\sqrt{(2\pi)^p |\boldsymbol{\Sigma}|}} \exp\left( -\frac{1}{2}(\mathbf{x} - \boldsymbol{\mu})\boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})^{\top} \right)
$$
where $|\boldsymbol{\Sigma}|$ denotes the determinant of $\boldsymbol{\Sigma}$. We write $\mathbf{X}\sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$. 

In R, the multivariate normal distribution is implemented, for instance, in the ``mvtnorm`` package. 
```{r}
library(mvtnorm)
```

For example, we can use ``dmvnorm()`` to compute the joint density function:
```{r}
dmvnorm(c(0, 0))
dmvnorm(c(0, 0, 1))
```
We can also provide several data point in the form of a matrix: 
```{r}
x <- matrix(c(c(0, 0), c(1, 1)), 2, 2, byrow = T)
x
dmvnorm(x)
```

Note that the default values for the mean vector is the vector of 0's and for the covariance matrix the identity matrix. We can change this values by providing the desired information in the respective arguments. 
```{r}
mu <- c(1, 1)
sigma <- matrix(c(4, 2, 2, 3), ncol = 2)
sigma
dmvnorm(c(0, 0), mean = mu, sigma = sigma)
```
Recall that for a random $\mathbf{X}$, its joint distribution function $F_{\mathbf{X}}$ is given by 
$$
F_{\mathbf{X}}(x_1, \dots, x_p) = \mathbb{P}( X_1 \leq x_1,\dots,  X_p \leq x_p) \,.
$$
The joint distribution function of $\mathbf{X}\sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ can be evaluated in R using the ``pmvnorm()`` function. For instance, we can compute 

$$
\mathbb{P}( X_1 \leq 1, X_2 \leq 4) \,.
$$
for $\mathbf{X}\sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ with $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$
```{r}
mu <- c(1, 1)
sigma <- matrix(c(4, 2, 2, 3), ncol = 2)
```
as follows
```{r}
pmvnorm(mean = mu, sigma = sigma, upper = c(1, 4))
```
``pmvnorm()`` also allow us to compute slightly more complicated probabilities. For example, to compute 
$$
\mathbb{P}( X_1 \leq 1, -1 < X_2 \leq 4) \,.
$$
we simply specify values in the ``lower`` argument of ``pmvnorm()``:
```{r}
pmvnorm(mean = mu, sigma = sigma, lower = c(-Inf, -1), upper = c(1, 4))
```

Simulation of multivariate normal models can be performed using the ``rmvnorm()`` function. For example, 

```{r}
x <- rmvnorm(n = 500, mean = c(1, 2), sigma = sigma)
head(x)
plot(x,
  main = "Simulated sample from multivariate normal",
  xlab = "X1", ylab = "X2"
)
```

:::{.remark}
The MLE's for $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$ are explicit and given by vector of sample means and the sample covariance matrix, which can be computed easy as
```{r}
colMeans(x)
var(x) * (1 - 1 / nrow(x))
```
We will cover in detail the estimation of more complex models in the next section. 
:::

Let us conclude the present section by recalling the close-form expression for the joint characteristic function of $\mathbf{X}\sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$

$$
\varphi_{\mathbf{X}} (\mathbf{t}) = \exp\left( i\mathbf{t}\boldsymbol{\mu}^{\top} - \frac{1}{2} \mathbf{t} \boldsymbol{\Sigma}\mathbf{t}^{\top}  \right) \,.
$$

### Copulas


A $p$-dimensional copula is a distribution function on $[0,1]^p$ with uniform marginals over $[0,1]$. In other words, for a $p$-dimensional random vector $\mathbf{U} = (U_1, \dots, U_p)$ on the unit cube, a copula $C$ is

$$
C(u_1, \dots, u_p) = \mathbb{P}(U_1 \leq u_1, \dots, U_p \leq u_p) \,.
$$

The importance of studying copulas for multivariate modeling is summarized by the following theorem, known as Sklar's Theorem.

::: {.theorem}
Let $F$ be a joint distribution function with marginals $F_1,\dots,F_p$. Then there exists a copula $C : [0,1]^p \to [0,1]$ such that, for all $x_1,\dots,x_p$ in $\bar{\mathbb{R}} = [-\infty,\infty]$,
\begin{equation}
F(x_1,\dots,x_p) = C(F_1(x_1),\dots,F_p(x_p))\,.
 (\#eq:sklar)
\end{equation}
Moreover, if the marginals are continuous, then $C$ is unique. Conversely, if $C$ is a copula and $F_1,\dots,F_p$ are univariate distribution functions, then the function $F$ defined in \@ref(eq:sklar) is a joint distribution function with margins $F_1,\dots,F_p$ .
:::


We now recall a fundamental proposition in probability theory. In particular, i) is key for performing stochastic simulation. 

:::{.proposition}
Let $F$ be a distribution function and $F^{\leftarrow}$ its generalized inverse. Then 

i) If $U$ is a standard uniform distributed random variable, that is, $U \sim U(0,1)$, then $\mathbb{P} (F^{\leftarrow}(U) \leq x) =  F(x)$. 

ii) If $Y \sim F$, with $F$ a continuous distribution function, then $F(Y) \sim U(0,1)$. 
:::

Thus, in the case of continuous marginals, Sklar's Theorem also suggests that we can find the copula of a given joint distribution. More specifically, if a random vector $\mathbf{X}$ has joint distribution function $F$ with continuous marginal distributions $F_1, \dots, F_p$, then the copula of $F$ (or $\mathbf{X}$) is the joint distribution function $C$ of the random vector $(F_1(X_1),\dots,F_d(X_p))$.


Copulas allow for modeling very different types of dependence structures. For instance, the independence case is retrieved by the *independence copula* given by 
$$
C_{ind}(u_1, \dots, u_p) = \prod_{i = 1}^p u_i \,.
$$
Indeed, Sklar's Theorem implies that r.v.'s (with continuous distributions) are independent if and only if their copula is the independence copula.

Another important property of copulas is their invariance under strictly increasing transformations of the marginals, as stated in the following proposition. 

::: {.proposition}
Let $(X_1,\dots, X_p )$ be a random vector with continuous marginals and copula $C$. Let $T_1,\dots, T_p$ be strictly increasing functions. Then $(T_1(X_1),\dots, T_p(X_p))$ also has copula $C$.
:::


As previously mentioned, copulas can be obtained from given distribution functions. In particular, we can derive copulas from several well-known multivariate distributions. These are commonly known as *implicit* copulas. For example, consider $\mathbf{X}\sim N(\boldsymbol{0}, \boldsymbol{P})$, where $\boldsymbol{0}$ is the $p$-dimensional vector of zeroes, and $\boldsymbol{P}$ is a $p \times p$ correlation matrix. Then, we define the *Gaussian* copula as the copula of $\mathbf{X}$. More specifically, the Gaussian copula is given by 

$$
C_{\boldsymbol{P}}^{Ga}(\mathbf{u}) = \mathbb{P}(\Phi(X_1) \leq u_1, \dots, \Phi(X_p) \leq u_p) = \boldsymbol{\Phi}_{\boldsymbol{P}} (\Phi^{-1}(u_1), \dots, \Phi^{-1}(u_p)) \,,
$$
where $\Phi$ denotes the distribution function of a (univariate) standard normal and $\boldsymbol{\Phi}_{\boldsymbol{P}}$ is the joint distribution function of $\mathbf{X}$. 
Note that considering more general $\mathbf{X}\sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$, leads to the same family of Gaussian copulas due to the property of invariance under increasing transformations.  

It turns out that the Gaussian copula falls into the more general family of *elliptical copulas*, which we will review next. 

#### Elliptical copulas {-}

Before introducing elliptic copulas, we define elliptical distributions. We say that a random vector $\mathbf{X}$ follows an elliptical distribution if its characteristic function $\varphi_{\mathbf{X}}$ is of the form 

$$
\varphi_{\mathbf{X}} (\mathbf{t}) = \exp( i\mathbf{t}\boldsymbol{\mu}^{\top}) \psi\left(\mathbf{t} \boldsymbol{\Sigma}\mathbf{t}^{\top}  \right) \,,
$$
for some vector $\boldsymbol{\mu}$ known as the location vector, some positive definite matrix $\boldsymbol{\Sigma}$ known as the dispersion matrix, and some function $\psi$. 
Note, for instance, that the multivariate normal distribution is a particular case with $\psi(t) = \exp(-\frac{1}{2}t)$. Another important example of elliptical distributions is the multivariate $t$ distribution. Recall that $\mathbf{X}$ is said to be multivariate $t$-distributed with $\nu$ degrees of freedom if its joint density function is given by
$$
f(\mathbf{x}) = \frac{\Gamma(\frac{1}{2}(v+ p))}{\Gamma(\frac{1}{2}v)\sqrt{(2\pi)^p |\boldsymbol{\Sigma}|}} \left( 1 + \frac{1}{v}(\mathbf{x} - \boldsymbol{\mu})\boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})^{\top} \right)^{-(v + p) /2} \,.
$$
We write $\mathbf{X} \sim t(\nu, \boldsymbol{\mu}, \boldsymbol{\Sigma})$. In this case, $\psi(t) = \hat{H}(-\frac{1}{2}t)$, where $\hat{H}$ is the Laplace transform of an (appropriate) Inverse Gamma distribution.


::: {.definition}
An *elliptical copula* is the implicit copula of an elliptical distribution. 
:::

:::{.remark}
Since copulas are invariant to increasing transformations of the marginals, elliptical copulas are typically defined in terms of the standardized dispersion matrix, or correlation matrix. 
:::

Now, let us review how to work with elliptical copulas in R. There are several packages to do so, for example, ``copula`` and ``fCopulae``. However, we will use ``copula`` here. 
```{r}
library(copula)
```
Elliptical copulas can be defined in R by using the ``ellipCopula()`` function, which creates objects of type ``copula``. Next, we present an example on how to define a Gaussian copula of dimension $2$ with correlation $0.4$:
```{r}
gaussian_cop <- ellipCopula(
  family = "normal", # Gaussian copula
  dim = 2, # Dimension of the copula
  dispstr = "ex", # Structure of the correlation matrix
  param = 0.4 # Correlation
)
gaussian_cop
```
The argument ``dispstr`` characterizes the structure of the correlation matrix. The available structures are "ex" for exchangeable, "ar1" for AR(1), "toep" for Toeplitz, and "un" for unstructured. For example, for dimension 3, the corresponding matrices are:

$$
\left( \begin{array}{ccc}
		1 &  \rho_1 & \rho_1\\
		\rho_1 &  1 & \rho_1  \\
		\rho_1 &  \rho_1 & 1  \\
	\end{array} \right) \,, \quad \left( \begin{array}{ccc}
		1 &  \rho_1 & \rho_1^2\\
		\rho_1 &  1 & \rho_1  \\
		\rho_1^2 &  \rho_1 & 1  \\
	\end{array} \right) \,, \quad \left( \begin{array}{ccc}
		1 &  \rho_1 & \rho_2\\
		\rho_1 &  1 & \rho_1  \\
		\rho_2 &  \rho_1 & 1  \\
	\end{array} \right) \,, \quad \mbox{and} \quad  \left( \begin{array}{ccc}
		1 &  \rho_1 & \rho_2\\
		\rho_1 &  1 & \rho_3  \\
		\rho_2 &  \rho_3 & 1  \\
	\end{array} \right) \,.
$$

For instance, following we use ``getSigma()`` to recover the correlation matrices of Gaussian copulas defined using the different values of ``dispstr``:
```{r}
getSigma(ellipCopula(
  family = "normal",
  dim = 3, dispstr = "ex", 
  param = 0.4
))

getSigma(ellipCopula(
  family = "normal",
  dim = 3, dispstr = "ar1", 
  param = 0.4
))

getSigma(ellipCopula(
  family = "normal",
  dim = 3, dispstr = "toep", 
  param = c(0.4, 0.5)
))

getSigma(ellipCopula(
  family = "normal",
  dim = 3, dispstr = "un", 
  param = c(0.4, 0.5, 0.2)
))
```

We can evaluate the density and distribution functions of a ``copula`` object via the ``dCopula()`` and ``pCopula()`` functions, respectively. For example, 

```{r}
dCopula(c(0.5, 0.2), gaussian_cop)
pCopula(c(0.5, 0.2), gaussian_cop)
```

Moreover, for 2-dimensional copulas, we can visualize the joint density function by using the ``contour()`` and ``persp()`` functions:
```{r}
par(mfrow = c(1, 2))
contour(gaussian_cop, dCopula)
persp(gaussian_cop, dCopula)
```

Finally, to generate random values from our copula object, we use ``rCopula()``.
```{r}
gauc_samp <- rCopula(5000, gaussian_cop)
head(gauc_samp)
plot(gauc_samp,
  main = "Simulation of Gaussian copula",
  xlab = "U1",
  ylab = "U2"
)
```

Now, let us exemplifying how to define a t-copula:

```{r}
t_cop <- ellipCopula(
  family = "t", 
  dim = 2,
  dispstr = "ex", 
  param = -0.4,
  df = 2
)
t_cop
```
Note that we can use the methods ``dCopula()``, ``pCopula()`` and ``rCopula()`` to compute the density and distribution functions and generate random values for this new object, respectively. For example, 

```{r}
tc_samp <- rCopula(5000, t_cop)
plot(tc_samp,
  main = "Simulation of t copula",
  xlab = "U1",
  ylab = "U2"
)
```


While elliptical copulas are implied by the well-known multivariate distribution functions of elliptical distributed random vectors, the copulas themselves do not have simple closed-forms. However, there are families of copulas that have simple closed-forms. These are often referred to as *explicit* copulas. Next, we review one of the most famous families of explicit copulas: the *Archemidiean* copulas.   

#### Archimedian copulas {-}

An Archimedian copula is characterized by a generator function $\phi : [0,1] \to [0,\infty]$ as

\begin{equation}
C(\mathbf{u}) = \phi^{-1}\left( \phi(u_1) + \cdots + \phi(u_p) \right)
(\#eq:arcop)
\end{equation}
where $\phi^{-1}$ is the inverse of the generator $\phi$. In order for \@ref(eq:arcop) to be a copula, a sufficient condition for the generator $\phi$ is that its inverse $\phi^{-1} :[0, \infty] \to [0,1]$ needs to be *completely monotonic*. In this family, the three most classical copulas are the Gumbel copula, the Frank copula and the Clayton copula. These copulas are constructed as follows:

**Gumbel copula.** The generator of this copula is $\phi(t) = (-\log(t))^{\alpha}$, $\alpha > 1$. Then

$$
C_{Gu}(\mathbf{u}) = \exp\left( - \left((-\log(u_1))^{\alpha} + \cdots + (-\log(u_p))^{\alpha} \right)^{1/\alpha} \right) \,.
$$

**Frank copula.**  In this case, the generator is $\phi(t) = -\log((\exp(-\alpha t) - 1) / (\exp(-\alpha ) - 1) )$, $\alpha \in \mathbb{R}$. Then

$$
C_{F}(\mathbf{u}) = -\frac{1}{\alpha}\log\left( 1 + \frac{(\exp(-\alpha u_1) - 1)  \cdots  (\exp(-\alpha u_p) - 1)}{\exp(-\alpha) - 1} \right) \,.
$$


**Clayton copula**  The generator of this copula is given by $\phi(t) = t ^{-\alpha} - 1$, $\alpha > 0$ (or $\alpha > -1$ for dimension 2). Then

$$
C_{C}(\mathbf{u}) =  \left(u_1^{-\alpha} + \cdots + u_p^{-\alpha} - p + 1 \right)^{-1/\alpha}  \,.
$$

::: {.remark}
Note that the three copulas above are uniparametric. 
::: 

In R, we can use the ``archmCopula()`` function to construct Archemidean copulas. For example, a 2-dimensional Gumbel copula of parameter $\alpha = 2.5$ is created as follows 
```{r}
gum_cop <- archmCopula(family = "gumbel", dim = 2, param = 2.5)
gum_cop
```

Again,  ``dCopula()``, ``pCopula()`` and ``rCopula()`` give access to the density and distribution functions and the random values generator, respectively.

For example, 
```{r}
par(mfrow = c(1, 3))
contour(archmCopula(family = "gumbel", dim = 2, param = 2),
  dCopula,
  main = "Gumbel copula",
  nlevels = 20
)
contour(archmCopula(family = "frank", dim = 2, param = 5.5),
  dCopula,
  main = "Frank copula",
  nlevels = 20
)
contour(archmCopula(family = "clayton", dim = 2, param = 2),
  dCopula,
  main = "Clayton copula",
  nlevels = 20
)
```

We will come back to some properties of these Archemedian copulas when we cover dependence measures. 


### Constructing multivariate distributions from copulas
Given a copula $C$ and marginals $F_1, \dots, F_p$, we now want to construct a multivariate distribution from these components. The ``copula`` package allows us to do precisely that via the ``mvdc()`` function, which creates objects of the type ``mvdc``. There are three main arguments for this function: ``copula`` which is a copula object, ``margins`` a character vector with the names of the marginals, and ``paramMargins`` which is a list with the parameters of the marginals. The use of this function is better illustrated with an example. Let us imagine that we want to create a bivariate distribution with normal distributed marginals $X_1 \sim N(1, 4)$ and $X_2\sim N(2, 2)$ and copula the Clayton copula with parameter $\alpha = 2$. First, we need the copula object:
```{r}
clay_cop <- archmCopula(family = "clayton", dim = 2, param = 2)
```
We can now define our ``mvdc`` object:
```{r}
my_mvd <- mvdc(
  copula = clay_cop,
  margins = c("norm", "norm"),
  paramMargins = list(list(mean = 1, sd = 2), list(mean = 2, sd = sqrt(2)))
)
my_mvd
```
We can now use ``dMvdc()`` to access the joint density of our multivariate model:
```{r}
dMvdc(c(1,2), my_mvd)
```
The joint distribution function can be evaluated using ``pMvdc()``:
```{r}
pMvdc(c(1,2), my_mvd)
```
Again, we can use ``contour()`` and ``persp()`` to visualize our distribution: 
```{r}
par(mfrow = c(1, 2))
contour(my_mvd, dMvdc, xlim = c(-4, 5), ylim = c(-2, 5)) # xlim and ylim must be given
persp(my_mvd, dMvdc, xlim = c(-4, 5), ylim = c(-2, 5))
```

Finally, random values from our multivariate model can be generated using ``rMvdc()`:
```{r}
my_mvdc_sample <- rMvdc(5000, my_mvd)
head(my_mvdc_sample)
plot(my_mvdc_sample,
  main = "Simulation of multivariate distribution",
  xlab = "X1",
  ylab = "X2"
)
```


::: {.remark}
We can pass user-defined distributions for the marginals as long as the density (``d``), distribution (``p``), and quantile  (``q``) functions are available. For example, if you name your distribution ``foo``, then you need the ``dfoo()``, ``pfoo()``, and ``qfoo()`` functions. 
:::


### Dependence measures

We now present some measures that assess the strength of dependence between random variables in different ways. We will limit ourselves to bivariate random vectors $(X_1,X_2)$ for presentation purposes.  

We first recall one of the most popular dependence measures: Pearson correlation. The correlation between $X_1$ and $X_2$, denoted by $\rho(X_1,X_2)$, is given by 

$$
\rho(X_1,X_2) = \frac{\mathbb{E}(X_1 X_2) - \mathbb{E}(X_1) \mathbb{E}(X_2) }{\sqrt{\mbox{Var}(X_1)\mbox{Var}(X_2)}}
$$
We know that if $X_1$ and $X_2$ are independent,then $\rho(X_1,X_2) =0$, but the converse is false.
If $|\rho(X_1,X_2)| = 1$ is equivalent to saying that $X_2 = a + b X_1$ for some $a \in \mathbb{R}$ and $b \neq0$. 


There are several pitfalls of using the correlation. For instance, we can have perfectly dependent random variables that exhibit relatively low correlation:

```{r}
set.seed(1)
x <- rexp(100)
y <- exp(x * 10)
cor(x, y)
```


Moreover, the correlation depends on the marginal distributions and may not even exist in some cases (we require finite second moments).

Hence, we now introduce other measures of dependence that depend only on the underlying copula of our multivariate model.

#### Rank correlation {-}

There are two crucial rank correlation measures: Kendall’s tau and Spearman’s rho. . The empirical estimators of rank correlations can be calculated by looking at the ranks of the data (i.e., the positions of the data), hence the name. 
We begin by defining Kendall’s tau, which can be understood as a measure of concordance for bivariate random vector. Recall that two point in $\mathbb{R}^2$, lets say $(x_1,x_2)$ and $(\tilde{x_1},\tilde{x}_2)$, are said to be *concordant* if $(x_1 - \tilde{x}_1)(x_2 - \tilde{x}_2)>0$ and *discordant* if $(x_1 - \tilde{x}_1)(x_2 - \tilde{x}_2)<0$. This motivates the following definition for random vectors $(X_1, X_2)$

:::{.definition}
For two random variables $X_1$ and $X_2$, Kendall's tau is given by
$$
\rho_\tau (X_1, X_2) = \mathbb{P}((X_1 - \tilde{X}_1)(X_2 - \tilde{X}_2)>0) - \mathbb{P}((X_1 - \tilde{X}_1)(X_2 - \tilde{X}_2)<0) \,.
$$
where $(\tilde{X}_1, \tilde{X}_2)$ is and independent copy of $(X_1, X_2)$
:::

In R, we can compute the empirical Kedall's tau by using ``method = "kendall"`` in ``cor()``:
```{r}
set.seed(1)
x <- rexp(100)
y <- exp(x * 10)
cor(x, y, method = "kendall")
```


The second way to measure rank correlation is with Spearma's rho. Although this measure can also be defined in terms of concordance and discordance for random pairs, we adopt the following definition:

:::{.definition}
For two random variables $X_1$ and $X_2$ with distribution functions $F_1$ and $F_2$, respectively, Spearma's rho is given by
$$
\rho_S (X_1, X_2) = \rho(F_1(X_1), F_2(X_2))\,.
$$
where $\rho$ denotes the Pearson correlation. 
:::

In R, the empirical Spearman's rho can be computed by changing ``method = "spearman"`` in ``cor()``:
```{r}
set.seed(1)
x <- rexp(100)
y <- exp(x * 10)
cor(x, y, method = "spearman")
```

As previously mentioned, an important feature of rank correlations is that they depend only on the (unique) copula of $(X_1, X_2)$. 

:::{.proposition}
Let $X_1$ and $X_2$ random variables with continuous marginal distribution functions and unique copula $C$. The rank correlations are given by
$$
\rho_\tau (X_1, X_2) = 4\int_{0}^1\int_{0}^1C(u_1,u_2)dC(u_1,u_2) - 1
$$
$$
\rho_S (X_1, X_2) = 12\int_{0}^1\int_{0}^1(C(u_1,u_2) - u_1u_2)du_1du_2
$$
:::


The ``copula`` package comes with the functions ``tau()`` and ``rho()`` to compute the rank correlations for a copula object. Let us give a example: 

```{r}
clay_cop <- archmCopula(family = "clayton", dim = 2, param = 2)
tau(clay_cop)
rho(clay_cop)
```



#### Coefficients of tail dependence {-}

The coefficients of tail dependence measure the strength of dependence in the tails of a bivariate distribution. These are defined as follows:

:::{.definition}
Let $X_1$ and $X_2$ be random variables with distribution functions $F_1$ and $F_2$, respectively. The *coefficient of upper tail dependence* $\lambda_u$ of $X_1$ and $X_2$ is 
$$
\lambda_u:= \lambda_u(X_1, X_2) = \lim_{q \to 1^{-}} \mathbb{P}(X_2>F_2^{\leftarrow}(q) \mid X_1>F_1^{\leftarrow}(q)) \,,
$$
provided a limit $\lambda_u \in [0,1]$ exists. If $\lambda_u \in (0,1]$, then we say that $X_1$ and $X_2$ show upper tail dependence. Alternatively, if $\lambda_u = 0$, then they are said to be asymptotically independent in the upper tail. Similarly, we define the *coefficient of lower tail dependence* $\lambda_l$ as 
$$
\lambda_l:= \lambda_u(X_1, X_2) = \lim_{q \to 0^{+}} \mathbb{P}(X_2\leq F_2^{\leftarrow}(q) \mid X_1\leq F_1^{\leftarrow}(q)) \,,
$$
provided a limit $\lambda_l \in [0,1]$ exists.
:::

When the distribution functions of $X_1$ and $X_2$ are continuous, the upper and lower tail dependence coefficients can be written (solely) in terms of the copula of the bivariate distribution. More specifically, we have that
$$
\lambda_u =\lim_{q \to 1^{-}} \frac{1 - 2q + C(q, q)}{1 - q} \quad \mbox{and} \quad \lambda_l =\lim_{q \to 0^{+}} \frac{C(q, q)}{q} \,.
$$

In R, we can compute these coefficients for a copula object using ``lambda()``. For example,

```{r}
lambda(clay_cop)
```



We conclude this section by providing Table \@ref(tab:depcop), which contains the closed-form formulas of some dependence measures for the copula models introduced so far.
``` {r, depcop, echo=FALSE}
d <- cbind(c("Gaussian", "t", "Gumbel", "Frank", "Clayton $(\\alpha>0)$"), c("$(2/ \\pi) \\arcsin (\\rho)$", "$(2/ \\pi) \\arcsin (\\rho)$", "$1 - 1/\\alpha$", "$1 - 4\\alpha^{-1}(1 - D_1(\\alpha))$", "$\\alpha / (\\alpha + 2)$"), c("0", "$2t_{\\nu+1}\\left( - \\sqrt{(\\nu + 1)(1 - \\rho) / (1 + \\rho)}\\right)$", "$2 - 2^{1/\\alpha}$", "0", "0"), c("0", "Same as $\\lambda_u$", "0", "0", "$2^{-1/\\alpha}$"))

knitr::kable(d, caption = "Kendall's tau and coefficients of tail dependence for some copulas. $D_1$ is the Debye function $D_1(\\theta) =\\theta^{-1}\\int_{0}^{\\theta}t / (\\exp(t) - 1)dt$ and $t_{\\nu +1}$ is the cdf of a univariate $t$ distribution with $\\nu +1$ degrees of freedom.", booktabs = TRUE, col.names = c("Copula", "$\\rho_{\\tau}$", "$\\lambda_{u}$", "$\\lambda_{l}$"), escape = FALSE)
```

### Fitting
Next, we will review two estimation methods for multivariate models based on copulas. 

#### Full maximum likelihood estimation {-}

Consider  $\mathbf{X} = (X_1, \dots, X_p)$ a random vector with joint distribution function $F$ which is specified by marginals with distribution functions $F_1 \dots, F_p$ and densities $f_1 \dots, f_p$, and a copula $C$ with density $c$. Furthermore, let $\boldsymbol{\beta}_j$ be the parameters of the $j$th marginal, that is, $f_j(\cdot, \boldsymbol{\beta}_j)$, $j =1, \dots, p$, $\boldsymbol{\alpha}$ be the parameters of the copula $C$, i.e., $C(\cdot, \boldsymbol{\alpha})$, and $\boldsymbol{\theta} = (\boldsymbol{\beta}_1, \dots,\boldsymbol{\beta}_p, \boldsymbol{\alpha})$. Next,
suppose that we have $n$ independent realizations  $\tilde{\mathbf{x}} = \{ \mathbf{x}_i = (x_{i1}, \dots, x_{ip}) : i = 1, \dots, n \}$ from $\mathbf{X}$. Then, the log-likelihood function is given by

$$
l(\boldsymbol{\theta}; \tilde{\mathbf{x}}) = \sum_{i = 1}^n \log(c(F_{1}(x_{i1}; \boldsymbol{\beta}_1), \dots, F_{p}(x_{ip}; \boldsymbol{\beta}_p); \boldsymbol{\alpha})) + \sum_{i = 1}^n \sum_{j= 1}^p \log(f_{j}(x_{ij}; \boldsymbol{\beta}_j)) \,.
$$
Thus, the maximum likelihood estimator $\hat{\boldsymbol{\theta}}$ of $\boldsymbol{\theta}$ is given by 
$$
\hat{\boldsymbol{\theta}} = {\arg\max}_{\boldsymbol{\theta} \in \boldsymbol{\Theta}} l(\boldsymbol{\theta}; \tilde{\mathbf{x}})\,.
$$
Let us now illustrate how to implement this procedure in R. Consider the following multivariate model: 

```{r}
my_mvd <- mvdc(
  copula = ellipCopula(family = "normal", param = 0.5),
  margins = c("gamma", "gamma"), 
  paramMargins = list(list(shape = 2,scale = 1), list(shape = 3, scale = 2))
)
```
Then, we simulate a sample from the model above 
```{r}
n <- 200
set.seed(1)
sim_dat <- rMvdc(n, my_mvd)
```
Firstly, we can make log-likelihood evaluations for certain set of parameters using the ``loglikMvdc()`` function. For example, using the original parameters of our model, we obtain:
```{r}
loglikMvdc(param = c(2, 1, 3, 2, 0.5), sim_dat, my_mvd)
```
Note that the parameters must be passed in vector form and ordered with the marginals first and the copula parameters last. Now, to perform MLE, we can use the ``fitMvd()`` function. In this case, we need to pass the data, an mvdc object with the structure that we want to fit, and the starting values for the optimization. For example
```{r}
mut_fit <- fitMvdc(sim_dat, my_mvd, start = c(1, 1, 1, 1, 0.1))
mut_fit
```
A useful characteristic of the implementation above is that we get an mvdc object with the estimated parameters as part of the output, which can be accessed using the ``@`` operator. Then, we can use all the methods for mvdc objects with our fitted multivariate model. For instance,

```{r, fig.align='center'}
persp(mut_fit@mvdc, dMvdc, xlim = c(0, 4), ylim = c(0, 6))
```




#### Two-step estimation {-}

When the dimension $p$ increases, the number of parameters to be estimated increases as well, making the optimization problem harder. In such a case, separating the estimation problem into fitting the marginals and the copula separately can significantly reduce the computational burden. This approach is often referred to as a two-stage parametric maximum likelihood method. More specifically, the method consists of estimating first the parameters of the marginals, that is, 
$$
\hat{\boldsymbol{\beta}}_j = {\arg\max}_{\boldsymbol{\beta}}  \sum_{i = 1}^n \log(f_{j}(x_{ij}; \boldsymbol{\beta})) \,, \quad \, j = 1,\dots, p \,.
$$
Next, we estimate the parameter $\boldsymbol{\alpha}$ of $C$, given the parameters of the marginals $\hat{\boldsymbol{\beta}}_j$, $j = 1,\dots, p$, by 


$$
\hat{\boldsymbol{\alpha}} = {\arg\max}_{\boldsymbol{\alpha}} \sum_{i = 1}^n \log(c(F_{1}(x_{i1}; \hat{\boldsymbol{\beta}}_1), \dots, F_{p}(x_{ip}; \hat{\boldsymbol{\beta}}_p); \boldsymbol{\alpha})) \,.
$$
In R, we can use the machinery introduced in Section \@ref(parinf) to fit the marginal distributions, and with those estimators at hand, then we can use the ``fitCopula()`` to find the MLE of the copula parameters. Let us now look to a specific example. First, we estimate the parameters of the marginals

```{r}
par1 <- fitdist(sim_dat[, 1], distr = "gamma")$estimate
par1
par2 <- fitdist(sim_dat[, 2], distr = "gamma")$estimate
par2
```
Next, using the estimators above, we find the evaluation points $(F_{1}(x_{i1}; \hat{\boldsymbol{\beta}}_1), F_{2}(x_{ip}; \hat{\boldsymbol{\beta}}_2))$ that will be used to estimate the copula parameters. 
```{r}
u_dat <- cbind(pgamma(sim_dat[, 1], par1[1], par1[2]),
    pgamma(sim_dat[, 2], par2[1], par2[2]))
```
Finally, we estimate the copula parameters using ``fitCopula()``
```{r}
cop_fit <- fitCopula(my_mvd@copula, u_dat, start = 0.1)
cop_fit
```
Thus, the estimated multivariate model would be the following: 
```{r}
mut_fit_2stage <- mvdc(
  copula = cop_fit@copula,
  margins = c("gamma", "gamma"),
  paramMargins = list(
    list(shape = unname(par1[1]), rate = unname(par1[2])),
    list(shape = unname(par2[1]), rate = unname(par2[2]))
  )
)
mut_fit_2stage
```
with corresponding log-likelihood
```{r}
loglikMvdc(
  param = c(par1, par2, cop_fit@copula@parameters),
  sim_dat,
  mut_fit_2stage
)
```



## Linear regression

Linear regression is one of the most commonly used statistical methods in practice. The main idea is to use explanatory variables (or covariates) to explain (and predict) a random variable of interest. More specifically, a linear regression model relates a response variable $Y$ to a $p$-dimensional vector $X = (X_1, \dots, X_p)$ of covariates via the identity

$$
\mathbb{E}\left[ Y \mid {X}\right] = \sum_{j = 1}^p X_j \beta_j =  X \boldsymbol{\beta}  \,,
$$
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_p)^{\top}$. It is common practice to add an *intercept* parameter $\beta_0$, so that the linear model becomes

$$
\mathbb{E}\left[ Y \mid {X}\right] = \beta_0 + X \boldsymbol{\beta}  \,.
$$
For notation convenience, the intercept is included among the other parameters, that is, $\boldsymbol{\beta} = (\beta_0, \beta_1, \dots, \beta_p)^{\top}$. Note that $\boldsymbol{\beta}$ is now a $(p +1)$ dimensional vector. Hence, we need to do an extra adjustment in order to write our regression model in terms of matrix products, and that is, to add the covariate $X_0 = 1$ to $X$. Thus,  our model can be written as

$$
\mathbb{E}\left[ Y \mid {X}\right] = X \boldsymbol{\beta} = \beta_0 + \sum_{j = 1}^p X_j \beta_j  \,.
$$
Now assume that we have a set of observations $(y_1, x_1), \dots, (y_n, x_n)$, where $x_i = (x_{i1}, \dots,x_{ip})$ are vectors of covariates, $i = 1,\dots, n$, and that we want to estimate $\boldsymbol{\beta}$ from this data. The most common method to do so is the method of *least squares*, which consists of finding the coefficients $\boldsymbol{\beta} = (\beta_1, \dots, \beta_p)$ that minimize the residual sum of squares

$$
RSS(\boldsymbol{\beta}) = \sum_{i = 1}^n (y_i - \beta_0 - \sum_{j = 1}^n x_{ij} \beta_j)^2 \,.
$$
Before showing how to minimize the above expression, we introduce further notation. Let $\mathbf{X}$ be the $n \times (p + 1)$ matrix with $i$-th row the vector $x_i$ (with $1$ in the first position) and let $\mathbf{y} = (y_1, \dots, y_n)^{\top}$. Then we can write the residual sum of squares as 
$$
RSS(\boldsymbol{\beta}) = (\mathbf{y} - \mathbf{X} \boldsymbol{\beta})^{\top}(\mathbf{y} - \mathbf{X} \boldsymbol{\beta})
$$
It is easy to see that the value of $\boldsymbol{\beta}$ that minimizes the expression above is given by 
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{\top}\mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y} \,.
$$
With the estimated parameters $\hat{\boldsymbol{\beta}}$ at hand, we can now compute the predict values $\hat{\mathbf{y}}$ for the input data as 

$$
\hat{\mathbf{y}} = \mathbf{X} \hat{\boldsymbol{\beta}}  \,.
$$
Moreover, we could use the above expression to predict values for new covariate information. 

In R, linear regression can be performed using the ``lm()`` function. Let us now look at a specific example, dealing with the *Prostate Cancer* data set available in the ``genridge`` R package. First, we need to load the data set into our working space:
```{r}
library(genridge)
data(prostate)
prostate <- prostate[, -10] # No relevant for the present analysis
head(prostate)
```
We aim to explain the variable ``lpsa`` (level of prostate-specific antigen) in terms of the other variables. The following is a scatter plot matrix showing every pairwise plot between the variables:
```{r}
plot(prostate)
```

One can observe, for example, that ``lcavol`` and ``lpc`` are positively correlated with ``lpsa``. Moreover, note that ``svi`` is a binary variable, and ``gleason`` is a categorical variable. 

We now fit a linear regression model using all our covariates:

```{r}
prost_lm <- lm(formula = lpsa ~ lcavol + lweight + age + lbph
  + svi + lcp + gleason + pgg45, data = prostate)
prost_lm
```
Using our linear model above, we can now computed the predicted values for the input data with the function ``predic()``: 
```{r}
pros_pred <- predict(prost_lm) 
head(pros_pred)
```

If we want to predict values of a new set of covariate values, these have to be passed into the ``predict()`` function in the form of a data frame with column names equal to those of the data frame used to create the linear mode. For example,
```{r}
xn <- matrix(c(1.35, 3.6, 64, 0.1, 0, -0.18, 6, 2.5), 1)
colnames(xn) <-  names(prostate[, -9])
xn <- as.data.frame(xn) # Covariates has to be in the form of a data.frame
xn
predict(prost_lm, xn) 
```


So far, we have made minimal assumptions regarding the linear model. However, we need further assumptions to draw inference about the parameters and the model. More specifically, we assume that our model is of the form

$$
 Y_i  =  X_i \boldsymbol{\beta} + e_i   \,, \quad i = 1,\dots, n \,.
$$

where the $e_i$, called are error terms or *residuals*, are uncorrelated with $\mathbb{E}[e_i] = 0$ and $Var[e_i] = \sigma^2$. Moreover, note that by using the vector notation $\mathbf{e} =(e_1, \dots, e_n)^{\top}$, we can rewrite the model as
$$
 \mathbf{Y}  =  \mathbf{X} \boldsymbol{\beta}+ \mathbf{e}.
$$
Under these assumptions, it is easy to see that the covariance matrix of the least squared estimator $\hat{\boldsymbol{\beta}}$ is given by

$$
Var(\hat{\boldsymbol{\beta}}) = \sigma^2 (\mathbf{X}^{\top}\mathbf{X})^{-1} 
$$
Note also that we can estimate the residuals as
$$
\hat{\mathbf{e}} = \mathbf{y} - \hat{\mathbf{y}}
$$
In R, one can compute the residuals of a linear model using the ``residuals()`` function. For instance, 

```{r}
res_pros <- residuals(prost_lm)
summary(res_pros)
```


It turns out that properties of the estimate residuals $\hat{\mathbf{e}}$, lead to the following unbiased estimator of the variance $\sigma^2$ 
$$
\hat{\sigma}^2 = \frac{1}{n - p - 1} ({\mathbf{y}}  - \hat{\mathbf{y}} )^{\top} ({\mathbf{y}}  - \hat{\mathbf{y}} ) = \frac{1}{n - p - 1} \sum_{i = 1}^n(y_i - \hat{y}_i)^2  \,.
$$
If we further assume that the residuals are Normal distributed with mean $0$ and variance $\sigma^2$, that is, $e_i \sim N(0, \sigma^2)$, $i =1 \dots, n$, then it is easy to show that 
$$
\hat{\boldsymbol{\beta}} \sim N({\boldsymbol{\beta}}, \sigma^2 (\mathbf{X}^{\top}\mathbf{X})^{-1} )
$$
and 

$$
(n - p - 1) \hat{\sigma}^2 \sim \sigma^2 \chi^{2}_{n - p - 1}
$$
a chi-squared distribution with $n - p -1$ degrees of freedom. Moreover, $\hat{\boldsymbol{\beta}}$ and $\hat{\sigma}^2$ are independent. These properties are particularly useful to perform hypothesis testing and create confidence intervals for the parameters $\beta_j$. 

To test the hypothesis that a coefficient of the linear model $\beta_j$ is equal to zero, i.e., $\beta_j = 0$, we consider the standardized coefficient or Z-score
$$
z_j = \frac{\hat{\beta_j}}{\hat{\sigma} \sqrt{v_j}} \,,
$$
where $v_j$ is the $j$-th diagonal element of $(\mathbf{X}^{\top}\mathbf{X})^{-1}$. Under the null hypothesis that $\beta_j = 0$, $z_j$ is $t_{n - p -1}$ distributed, and hence large (absolute) values of $z_j$ will lead to the rejection of this null hypothesis. In R, we can use ``summary()`` to access the values of these statistics (*t-statistic* column):

```{r}
summary(prost_lm)
```

R also provides the *p-value*s associated to these tests (column *Pr(>|t|)*). Recall that a p-value is the probability of obtaining test results at least as "extreme" as the results observed, assuming that the null hypothesis is correct. In other words, the smaller the p-value, the stronger the evidence that you should reject the null hypothesis. A nice feature of R is that marks with ``*`` and ``.`` the parameters with p-values lower than commonly used significance levels. 


It is often the case that we need to test for significance of groups of coefficients simultaneously. For that purpose, we can use the F statistics, defined as
$$
F = \frac{(RSS_0 - RSS_1) / ( p_1 - p_0)}{RSS_1 / (n - p_1 - 1)} \,,
$$
where $RSS_1$ is the residual sum of squares for the fit of the bigger model with $p_1 + 1$ parameters, and $RSS_0$ the corresponding value for the (nested) smaller model with $p_0 + 1$ parameters. Under or Gaussian assumption on the residuals, and that the smaller model is correct, the F statistics is $F_{p_1 - p_0, n - p_1 - 1}$ distributed. In the output of the ``summary()`` function above, we have an F statistics (and corresponding p-value) that compares with a model with all coefficients equal to zero (i.e., test the null hypothesis: "All coefficients are zero"). 

Finally, note that  ``summary()`` displays additional information such as the *Multiple R-squared*. This measures the proportion of the variance in the response variable that the model can explain, and it varies from 0 to 1. Thus, we should aim for a model with multiple R-squared as close to 1 as possible. The last piece of information is the *Adjusted R-squared*, which is simply an adjustment to the multiple R-squared that considers the data size and the number of covariates. Hence, the adjusted R-squared is recommended to be used over the (conventional) multiple R-squared.



Considering all the information described above, let us now consider a smaller model that consists only of the covariates ``lcavol``,  ``lweight``,  ``lbph``, ``svi`` and without an intercept. 


```{r}
prost_lm2 <- lm(formula = lpsa ~ lcavol + lweight
  + lbph + svi - 1, data = prostate)
summary(prost_lm2)
```

We can see that for the covariates selected, we can reject the null hypothesis $\beta_j = 0$, and we obtain a larger adjusted R-squared. All good indicators that our model is better.

Furthermore, we can test if "the smaller model is correct" using the F statistics, which can be accessed in R via the ``anova()`` function:

```{r}
anova(prost_lm, prost_lm2)
```
Hence, we cannot reject the null hypothesis that "the smaller model is correct". 



We conclude our discussion on linear models with a discussion on how to construct confidence intervals for $\beta_j$. It is easy to see that under the Gaussian assumption on the residuals, a $(1 - \alpha)$ confidence interval for $\beta_j$ is given by
$$
\hat{\beta}_j \pm z_{1 - \alpha /2} v_j^{1/2} \hat{\sigma} \,,
$$
where $z_{1 - \alpha /2}$ is the $(1 - \alpha /2)$ quantile of a standard normal distribution. We can compute confidence intervals in R via the ``confint()`` function. For instace, for our last model we have
```{r}
confint(prost_lm2)
```
Corresponding to 95% confidence intervals for the parameters $\beta_j$. 

